---
title: 'ggplot2 intro exercise'
author: "Mike Frank"
date: "November 10, 2021"
output: html_document
---
  
This is an in-class exercise on using `ggplot2`.

Note, that this example is from the_grammar.R on [http://had.co.nz/ggplot2](). I've adapted this for psych 251 purposes

First install and load the package. It's part of the "core tidvyerse".

```{r}
library(tidyverse)
```

# Exploring ggplot2 using `qplot`

We'll start by using `qplot`. `qplot` is the easy interface, meant to replace `plot`. You can give it simple `qplot(x,y)` examples, or slightly more complex examples like `qplot(x, y, col=grp, data=d)` (`col` sets the color, `data` pulls from a dataframe). 

We're going to be using the `diamonds` dataset. This is a set of measurements of diamonds, along with their price etc.

```{r}
head(diamonds)
```

Try making a scatter plot of diamonds' carats (`diamonds$carat`) plotted by their price (`diamonds$price`) using `qplot`. 

```{r}
qplot(diamonds$carat, diamonds$price)
```

Now make a histogram of prices.

```{r}
qplot(diamonds$price)
```


Scatter plots are trivial to make and also easy to add features to. Modify this plot so that it uses the dataframe `diamonds` rather than working from variables in the general namespace (good to get away from retyping `diamonds$` every time you reference a variable). 

```{r}
qplot(x=carat, y=price, data=diamonds)
```

Try adding clarity and cut, using shape and color as your visual variables (these are just arguments to the qplot function). 

```{r}
qplot(data = diamonds,
      x=shape, )
```


# More complex with `ggplot`

`ggplot` is just a way of building `qplot` calls up more systematically. It's
sometimes easier to use and sometimes a bit more complicated. What I want to show off here is the functionality of being able to build up complex plots with multiple elements. You can actually do this using `qplot` pretty easily, but there are a few things that are hard to do. 

`ggplot` is the basic call, where you specify A) a dataframe and B) an aesthetic (`aes`) mapping from variables in the plot space to variables in the dataset. 

```{r}
p <- ggplot(diamonds, aes(x=carat, y=price)) # first you set the aesthetic and dataset
p <- p + geom_point() # then you add geoms
p <- p + geom_point(aes(colour = carat)) # and you can keep doing this to add layers to the plot
p
```

Try writing this as a single set of additions (e.g. one line of R code, though you can put in linebreaks). This is the most common workflow for me. 


```{r}
ggplot(...) + 
  geom_point(...)
```


# Facets

Let's try out another version of the `qplot` example above. Rewrite the last qplot example with ggplot.

```{r}
...
```

One of the primary benefits of `ggplot2` is the use of facets - also known as small multiples in the Tufte vocabulary. That last plot was probably hard to read. Facets could make it better. Try adding `facet_grid(x ~ y)`.  `x ~ y` means row facets are by `x`, column facets by `y`. 

```{r}
ggplot(diamonds, 
      aes(x = carat, y = price, )) + 
  geom_point(aes(color=clarity)) + 
  facet_grid(cut ~ color)
```

But facets can also get overwhelming. Try to strike a good balance between color, shape, and faceting.

HINT: `facet_grid(. ~ x)` puts x on the columns, but `facet_wrap(~ x)` (no dot) *wraps* the facets. 

```{r}
ggplot(diamonds, 
      aes(x = carat, y = price)) +
  geom_point(aes( color = clarity)) + 
  geom_smooth(method="lm") + 
  facet_wrap(~ cut)
```


# Geoms

As you've seen above, the basic unit of a ggplot plot is a "geom" - a mapping between data (via an "aesthetic") and a particular geometric configuration on coordinate axes. 

Let's try adding some geoms and manipulating their parameters. One combo I really like is a scatterplot with a smoothing layer (`geom_smooth`). Try adding one onto this plot. 

```{r}
ggplot(diamonds, aes(x=carat, y=price)) +
  geom_point(shape = ".") +
  geom_smooth() + 
  facet_grid(cut ~ clarity) 
```


CHALLENGE: You could also try starting with a histogram and adding densities. (`geom_density`), Try [this link](https://stackoverflow.com/questions/5688082/ggplot2-overlay-histogram-with-density-curve).

```{r}

```

# Themes and plot cleanup

I like a slightly cleaner look to my plots. Luckily, ggplot allows you to add "themes" to your plots. Try doing the same plot but adding `+ theme_bw()` or `+ theme_classic()`. Different themes work better for different applications, in my experience. 

You can also try different color scales. Experiment with `scale_color_...` - try writing that and hitting TAB for autocomplete options. Check out the help on this. 

You can also try transparency/different point sizes to clean up scatterplots. Try `alpha = .1` or `pch = "."` to make transparent or small points. 

Finally, don't forget to "fix the axis labels"! 

Here's a somewhat ugly plot - see if you can make it look awesome. 

```{r}
ggplot(diamonds, aes(x = carat, y = price, 
                     col = cut)) + 
  geom_point(alpha=.1) + 
  facet_wrap(~clarity) +
  theme_minimal()
```

Now you try it:

```{r}
```